Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-02-10 LaTeX/Turnover.tex
figure legends and other changes

We note two other ways in which sex determination has been shown to relate to zygotic sex ratios.
Firstly, non-Fisherian female-biased sex ratios can be favoured when there is local mate competition, in which brothers compete for mates \citep{Hamilton:1967ts}. 
Therefore, with local mate competition, feminizing mutations can spread because they bias the sex ratio towards females \citep{Wilson:1981vm,Vuillleumier:2007bh}. 
Secondly, environmental conditions (e.g., maternal condition, mate quality, age, or host size) can differentially affect the fitness of males versus females such that the optimal allocation to males/females depends on the environment \citep{Trivers:1973wb,Charnov:1977tx,Charnov:1982wg}. 
In such cases, flexible sex determination systems may evolve in order to allow the zygotic sex ratio to be determined in a way that depends on the environment \citep{Charnov:1977tx,Werren:1984tl,Pen:2010kk}. 
In this study, we do not consider environmental condition dependence or local mate competition \citep[reviewed in][]{Charnov:1982wg,West:2009we}.

%\textcolor{red}{We add haploid selection:}
Here, we use mathematical models to find the conditions under which new sex determination systems are favoured when loci experience haploid selection. 
Haploid genotypes at many loci experience selection during gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-medelian) segregation of genotypes during gamete production and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool; the term `haploid selection' encompasses both processes. 
Meiotic drive generally occurs either during the production of male or female gametes only \citep{Ubeda:2005gw,Lindholm:2016cw}.
Because there are typically more pollen/sperm than required for fertilization, gametic competition is also typically sex specific, occurring primarily among male gametes.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller (but non-negligible) proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.

@@

Fisherian sex ratio selection is sometimes considered in terms of balancing parental investment in male versus female offspring \textcolor{red}{(citation)}. 
In addition, under environmental sex-determination, the proportion of males/females is sometimes controlled by the mother, e.g., the proportion of eggs laid in warm versus cold environments could determine the sex ratio of offspring. 
We therefore also considered the invasion of a neo-sex-determining allele ($m$) in a model in which mothers that have at least one $m$ allele produce $k$ daughters and $(1-k)$ males. 
As with offspring-controlled ESD, for all $k\in\{0,1/2,1\}$, we find that invasion into an ancestrally XY system is the same as invasion into an ancestrally ZW system (at least up to order $\epsilon^3$), implying transitions between genetic sex determination and maternally controlled environmental sex determination are not driven by Fisherian sex ratio selection caused to biased zygotic sex ratios.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Heterogametic transitions with/without sex ratio change
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{Combination_Turnover}\\
\caption{
Heterogametic transitions from XY to ZW sex determination (neo-W frequency shown by black lines, panels A and B) or from ZW to XY (neo-Y frequency shown by black lines, panels C and D) occurs similarly regardless of sex ratio biases present before (B versus D) or after (C versus A, dashed lines show male frequency). 
During the invasion of a neo-ZW sex determination system (A and B), the ancestral Y fixes in both males and females (blue and red lines). 
Similarly, the ancestral W allele fixes in males and females (blue and red lines) during a ZW to XY transition. 
In this plot, there is no gametic competition ($t^\female=t^\male=0$) and meiotic drive occurs during male meiosis only ($\alpha^\female_{\Delta}=0$, $\alpha^\Hermaphrodite_{\Delta}=-1/5$). Therefore, sex ratio biases can only arise when the \textbf{A} locus is linked to an XY sex-determining locus.
In panels A and C, the neo-sex-determining locus is more closely linked to the \textbf{A} locus than the ancestral sex-detemining region ($r=1/2$, $R=1/20$) such that a neo-Y can caused biased sex ratios (panel C). 
Unlike with diploid selection alone, when there is haploid selection (in this case meiotic drive), neo-sex-determining loci that are less closely linked to the \textbf{A} locus can also spread (panels B and D, $r=1/20$, $R=1/2$), see equation \eqref{eq:lambda_neoW} and Figure \ref{fig:Combination_Centimorgans}B. 
These transitions are unusual because linkage generally allows favourable associations to arise via selection and the new sex determination systems in B and D have looser linkage. 
Thus, diploid mean fitness decreases over the course of the transitions in B and D, see Figure \ref{fig:Combination_MeanFit}.
However, the mean fitness of females increases during the spread of dominant neo-W alleles and the mean fitness of males increases during the spread of dominant neo-Y alleles, Figure \ref{fig:Combination_MeanFit}. 
In this plot there are no sex differences in selection and an equilirbium is maintained because selection in diploids opposes meiotic drive, $s^\female =s^\male = 1/5$, $h^\female = h^\male = 7/10$.
\textcolor{red}{Aethetic adjustments: Could add titles to the columns/rows: neo-W for row 1, neo-Y for row 3, $r=0.5$, $R=0.05$ for column 1 and $r=0.05$, $R=0.5$ for column 2. Could adjust padding (too much whitespace where there is no axis label). It also seems could increase ratio of font size relative to plot size to make figure more compact. Could make sex ratio biases more extreme by reducing the $r$ in A and C and reducing $R$ in B and D. Matt - could you uncomment the line legends in the Mathematica file (function not included in my Mathematica version).}
}
\label{fig:Combination_Turnover}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Chromosomal Positions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{Combination_Centimorgan}\\
\caption{
The invasion fitness of a neo-sex-determining locus with various 
\textcolor{red}{A sexual antagonism (no haploid selection), B drive (no gametic competition), equal selection in sexes ($s^\female=s^\male$), C \& D Pollen/Sperm competition only (no drive). C allele favoured in pollen/sperm competition selected against less in males ($t<0$, $s^\female, s^\male >0$, $s^\female<s^\female$). D allele favoured in pollen/sperm competition selected against more in males than females ($t<0$, $s^\female, s^\male >0$, $s^\female>s^\female$). I suspect that panel C has a region where no equilibrium is maintained (CHECK! Maybe include different parameters here). Currently use different parameters for B than using in figure \ref{fig:Combination_Turnover} (selection/drive twice as strong in turnover figure)}
}
\label{fig:Combination_Centimorgans}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

@@

\section*{Supplementary Figures}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Changes in mean fitness
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{Combination_MeanFit}\\
\caption{
\textcolor{red}{Could add titles to the columns/rows: neo-W for row 1, neo-Y for row 3, $r=0.5$, $R=0.05$ for column 1 and $r=0.05$, $R=0.5$ for column 2. \& possibly adjust padding (too much whitespace?). Matt - could you uncomment the line legends in the Mathematica file (function not included in my Mathematica version).}
}
\label{fig:Combination_MeanFit}
\end{figure}

>>>>

We note two other ways in which sex determination has been shown to relate to zygotic sex ratios.
Firstly, female-biased sex ratios can be favoured when there is local mate competition, in which brothers compete for mates \citep{Hamilton:1967ts}. 
Therefore, with local mate competition, feminizing mutations can spread because they bias the sex ratio towards females \citep{Wilson:1981vm,Vuillleumier:2007bh}. 
Secondly, environmental conditions (e.g., maternal condition, mate quality, age, or host size) can differentially affect the fitness of males versus females such that the optimal allocation to males/females depends on the environment \citep{Trivers:1973wb,Charnov:1977tx,Charnov:1982wg}. 
In such cases, flexible sex determination systems may evolve in order to allow the zygotic sex ratio to be determined in a way that depends on the environment \citep{Charnov:1977tx,Werren:1984tl,Pen:2010kk}. 
In this study, we do not consider environmental condition dependence or local mate competition \citep[reviewed in][]{Charnov:1982wg,Bull:1983vi,West:2009we}.

%\textcolor{red}{We add haploid selection:}
Here, we use mathematical models to find the conditions under which new sex determination systems are favoured when loci experience haploid selection. 
Haploid genotypes at many loci experience selection during gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-medelian) segregation of genotypes during gamete production and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool; the term `haploid selection' encompasses both processes. 
Meiotic drive generally occurs either during the production of male or female gametes only \citep{Ubeda:2005gw,Lindholm:2016cw}.
Because there are typically more pollen/sperm than required for fertilization, gametic competition is also typically sex specific, occurring primarily among male gametes.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A much smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.

@@

Fisherian sex ratio selection is sometimes considered in terms of balancing parental investment in male versus female offspring \citep{Charnov:1982wg}.
In addition, under environmental sex-determination, the proportion of males/females is sometimes controlled by the mother, e.g., the proportion of eggs laid in warm versus cold environments could determine the sex ratio of offspring. 
We therefore also considered the invasion of a neo-sex-determining allele ($m$) in a model in which mothers that have at least one $m$ allele produce $k$ daughters and $(1-k)$ males. 
As with offspring-controlled ESD, for all $k\in\{0,1/2,1\}$, we find that invasion into an ancestrally XY system is the same as invasion into an ancestrally ZW system (at least up to order $\epsilon^3$), implying transitions between genetic sex determination and maternally controlled environmental sex determination are not driven by Fisherian sex ratio selection on biased zygotic sex ratios.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Heterogametic transitions with/without sex ratio change
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{Combination_Turnover}\\
\caption{
Heterogametic transitions from XY to ZW sex determination (neo-W frequency shown by black lines, panels A and B) or from ZW to XY (neo-Y frequency shown by black lines, panels C and D) occur similarly regardless of sex ratio biases present before (B versus D) or after (C versus A, dashed lines show male frequency). 
During invasion by a neo-ZW sex determination system (A and B), the ancestral Y fixes in both males and females (blue and red lines). 
Similarly, the ancestral W allele fixes in males and females (blue and red lines) during a ZW to XY transition. 
In this plot, there is no gametic competition ($t^\female=t^\male=0$) and meiotic drive occurs during male meiosis only ($\alpha^\female_{\Delta}=0$, $\alpha^\Hermaphrodite_{\Delta}=-1/5$). Therefore, sex ratio biases can only arise when the \textbf{A} locus is linked to an XY sex-determining locus.
In panels A and C, the neo-sex-determining locus is more closely linked to the \textbf{A} locus than the ancestral sex-detemining region ($r=1/2$, $R=1/20$) such that a neo-Y can caused biased sex ratios (panel C).
In panels B and D, the ancestral sex-determining locus is more closely linked to the \textbf{A} locus than the neo-sex-detemining locus ($r=1/20$, $R=1/2$). 
Therefore, an ancestral XY sex determination can have a biased zygotic sex ratio that becomes unbiased after an unlinked neo-W invades (B). 
However, in panel D, a unlinked neo-Y invades an ancestral ZW sex determination system in a similar manner but no biases to the zygotic sex ratio occur. 
With diploid selection alone, neo-sex-determining loci do not spread if they are less closely linked to the \textbf{A} locus than the ancestral sex-determining locus (see equation \eqref{eq:lambda_neoW} and Figure \ref{fig:Combination_Centimorgans}A). 
In this plot there are no sex differences in selection and an equilibrium is maintained because selection in diploids opposes meiotic drive, $s^\female =s^\male = 1/5$, $h^\female = h^\male = 7/10$.
\textcolor{red}{Aethetic adjustments: Could add titles to the columns/rows: neo-W for row 1, neo-Y for row 3, $r=0.5$, $R=0.05$ for column 1 and $r=0.05$, $R=0.5$ for column 2. Could adjust padding (too much whitespace where there is no axis label). It also seems could increase ratio of font size relative to plot size to make figure more compact. Matt - could you uncomment the line legends in the Mathematica file (function not included in my Mathematica version).}
}
\label{fig:Combination_Turnover}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Changes in mean fitness
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{Combination_MeanFit}\\
\caption{
Here, we plot how male mean fitness (blue lines), female mean fitness (red lines), and population mean fitness (male mean fitness plus female mean fitness, black lines) changes during the transitions between sex-determination systems shown in Figure \ref{fig:Combination_Turnover}. 
Here we multiply male mean fitness and female mean fitness by two so that we can show it on the same scale as population mean fitness. 
The mean fitness of females increases during the spread of neo-W alleles (A and B) and the mean fitness of males increases during the spread of neo-Y alleles (C and D). 
However, when a neo-sex determining system evolves that is less closely linked to a locus under selection (B and D), population mean fitness decreases. 
\textcolor{red}{
Could add titles to the columns/rows: neo-W for row 1, neo-Y for row 3, $r=0.5$, $R=0.05$ for column 1 and $r=0.05$, $R=0.5$ for column 2. \& possibly adjust padding (too much whitespace?). Matt - could you uncomment the line legends in the Mathematica file (function not included in my Mathematica version).}
}
\label{fig:Combination_MeanFit}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Chromosomal Positions
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{Combination_Centimorgan}\\
\caption{
The invasion fitness of a neo-W allele plotted against the relative location of a locus under direct selection, \textbf{A}, for various selective regimes. 
We assume that the ancestral sex-determining locus is located at -0.25, the novel sex-determining locus is located at 0.25 and that there is a polymorphism at the \textbf{A} locus maintained by selection.
We used Haldane's map function \citep[Equation 3 in ][]{Haldane1919} to convert from map distance (centiMorgans) to the probability of a cross-over event. 
In A, there is no haploid selection ($t^\Hermaphrodite=\alpha^\Hermaphrodite_{\Delta}=0$) and selection in diploids is sexually antagonistic \citep[following][]{vanDoorn:2010hu}, in which case a neo-W can only invade if it is more closely linked to the selected locus ($s^\female=1/10$, $h^\female=7/10$ $s^\male=-1/10$, $h^\male=3/10$).
In B-D we include haploid selection and assume that selection in diploids is not sexually-antagonistic ($s^\female s^\male>0$). 
A polymorphism can then be maintained by opposing selection between the haploid and diploid phases. 
In B, there is drive in favour of the $a$ allele in males ($\alpha^\male_{\Delta}=-1/10$), no female meiotic drive or gametic competition, $t^\Hermaphrodite=\alpha^\female_{\Delta}=0$), and equal selection in diploid sexes ($s^\female=s^\male=1/10$, $h^\female=h^\male=7/10$). In this case, a neo-W can invade even when the selected locus is more closely linked to the ancestral sex determining locus (see Table \ref{tab:specialcases} and Figure \ref{fig:Combination_Turnover}). 
In C and D, there is gametic competition among male gametes only (favouring $a$, $t^\male=-1/10$) and no meiotic drive or gametic competition in females ($t^\female=\alpha^\Hermaphrodite_{\Delta}=0$). 
In this case, if the selected locus is closely linked to the ancestral sex-determining locus, the neo-W only invades when the allele favoured in pollen/sperm competition selected against more in males than females (D versus C, see Table \ref{tab:specialcases}). In C, $s^\female=3/20$, and $s^\male=1/20$, whereas in D, $s^\female=1/20$, and $s^\male=3/20$. In both C and D, $h^\female=h^\male=7/10$.
\textcolor{red}{I suspect that panel C has a region where no equilibrium is maintained (CHECK! Maybe include different parameters here or remove the part when no equilibrium). Currently use different parameters for B than using in figure \ref{fig:Combination_Turnover} (selection/drive twice as strong in turnover figure). This plot would also benefit from titles giving, e.g., ``sexually-antagonistic selection, $s^\female s^\male<0$'' for A, ``male meiotic drive, $s^\female s^\male>0$'' for B}
}
\label{fig:Combination_Centimorgans}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

@@

%\section*{Supplementary Figures}
